### 2024-02-22 misc/expo/entropy.tex
PFR progress

\def\eps{\epsilon}
\def\FF{{\bf F}}
\def\bar#1{\overline{#1}}
\def\hat#1{\widehat{#1}}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
\def\Bohr{\op{\rm Bohr}}
\def\Eta{\op{\bf H}}
\def\Etaseven{\op{\sevenbf H}}
\def\II{\op{\bf I}}
\def\dd{\op{\bf d}}
\def\ee{\op{\bf e}}
\def\given{\mathbin{|}}
\def\argmax{\limitop{arg$\;$max}}
\def\advthm{\the\sectcount.\the\thmcount\global\advance \thmcount by 1}
\def\advsect{\global\advance\sectcount by 1\section\the\sectcount\global\thmcount=1. }
\sectcount=0

@@

\maketitle{Entropy and additive combinatorics}{by}{Marcel K. Goh}{15 February 2024}

@@

We shall take it on faith that there really exists a function on random variables satisfying
these axioms.
In fact, the axioms only define entropy
up to a multiplicative constant, so we shall add the following axiom.
(It is very possible you have met this function somewhere on your travels, but you will not find it anywhere in
these notes.)

@@

\edef\lemmaxprob{\the\sectcount.\the\thmcount}
\proclaim Proposition \advthm. Let $X$ be a discrete random variable supported on a set $A$ and let
$$a^* = \argmax_{a\in A} \pr\{X = a\}.$$
Then
$$\pr\{X = a^*\} \ge 2^{-\Etaseven\{X\}}.$$

@@

\medskip\boldlabel Mutual information.
For random variables $X$ and $Y$, the {\it mutual information} $\II\{X : Y\}$ is defined
by the equivalent formulas
$$\eqalign{
\II\{ X : Y\} &= \Eta\{X\} + \Eta\{Y\} - \Eta\{X,Y\} \cr
&= \Eta\{X\} - \Eta\{X\given Y\} \cr
&= \Eta\{Y\} - \Eta\{Y\given X\}. \cr
}$$
It measures, roughly speaking, how much information one can get from one variable
by looking at the other one. From the formula it is clear that $\II\{X : Y\}  = 0$
if $X$ and $Y$ are independent. In general, we still have the inequality $\II\{X : Y\} \ge 0$,
which a corollary of the following lemma, which expresses the intuitive fact that conditioning
cannot increase the entropy of a random variable.
The proof, which is due to C.~West, uses an argument similar to the one
we used to prove Lemma~{\lemmaxprob}.

@@

The general case follows from the continuity axiom and the fact that any discrete random
variable can be approximated by discrete random variables with finite support and rational
probabilities.\slug

@@

\proclaim Corollary \advthm. If $X$ and $Y$ are independent, then
$$\max\bigl( \Eta\{X\}, \Eta\{Y\} \bigr) \le \Eta\{ X \pm Y\}.$$

@@

The entropic analogue of the Ruzsa distance is defined as follows.
For finitely supported random variables $X$ and $Y$ taking values in the same abelian group,
we let $X'$ and $Y'$ be independent copies of $X$
and $Y$, respectively, and define the {\it entropic Ruzsa distance} by
$$\dd\{X,Y\} = \Eta\{X'-Y'\} - {\Eta\{X'\} \over 2} - {\Eta\{Y'\}\over 2}.$$
This definition only depends on the individual distributions of $X$ and $Y$ and does not require them to
have the same sample space. Once again, we don't necessarily have $\dd\{X,X\} = 0$, but we do have
the triangle inequality, which shall now prove.

@@

If $A$ and $B$ are subsets of the same abelian group such that $|A+B|$ is much smaller than $|A|\cdot|B|$,
then it stands to reason that there must be a lot of redundancy in $A+B$; that is, many elements of $A+B$ can
be expressed as $a+b$ in lots of different ways. To capture this notion, we can define the
{\it additive energy} between two sets $A$ and $B$ to be
$$E(A,B) = \bigl| \bigl\{ (a,a',b,b') \in A\times A\times B\times B : a + b = a'+b'\bigr\}\bigr\}.$$

@@

\proclaim Proposition \advthm. Let $A$ be a finite subset of an abelian group $G$. If $|A+A| = |A|$,
then $A$ is a coset of a subgroup $H$.

@@

If $x,y\in H$, then $A+x = A$ and $A+y=A$, so adding $y$ to both sides of the second identity we obtain
$A-y = A$, and adding $x$ to both sides of this, we get
$$A + x-y = A+x = A,$$
so $x-y\in H$.
This along with the fact that $0\in H$ shows that $H$ is a subgroup of $G$.

@@

The proof, which is almost entirely information-theoretic in nature, will be our main concern for the
remainder of these notes.

>>>>

\def\eps{\epsilon}
\def\FF{{\bf F}}
\def\bar#1{\overline{#1}}
\def\hat#1{\widehat{#1}}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
\def\Bohr{\op{\rm Bohr}}
\def\Eta{\op{\bf H}}
\def\Etaseven{\op{\sevenbf H}}
\def\II{\op{\bf I}}
\def\dd{\op{\bf d}}
\def\ee{\op{\bf e}}
\def\P{{\cal P}}
\def\given{\mathbin{|}}
\def\argmax{\limitop{\rm arg$\,$max}}
\def\dTV{d_{\rm TV}}
\def\advthm{\the\sectcount.\the\thmcount\global\advance \thmcount by 1}
\def\advsect{\global\advance\sectcount by 1\section\the\sectcount\global\thmcount=1. }
\sectcount=0

@@

\maketitle{Entropy and additive combinatorics}{}{Marcel K. Goh}{15 February 2024}

@@

The continuity axiom presupposes a topology on the set of all distributions on the target
space of $X$. This topology will only be important here and there, so we will not give a complete
description of it now, but simply introduce the relevant properties as needed later on.

We shall take it on faith that there really exists a function on random variables satisfying
these axioms.
In fact, the axioms only define entropy
up to a multiplicative constant, so we shall add the following axiom.
(It is very possible you have met the formula for entropy
somewhere on your travels, but you will not find it anywhere in
these notes.)

@@

\edef\propmaxprob{\the\sectcount.\the\thmcount}
\proclaim Proposition \advthm. Let $X$ be a discrete random variable supported on a set $A$ and let
$$a^* = \argmax_{a\in A} \pr\{X = a\}.$$
Then
$$\pr\{X = a^*\} \ge 2^{-\Etaseven\{X\}}.$$

@@

\medskip\boldlabel Mutual information.
For random variables $X$ and $Y$, the {\it mutual information} $\II\{X : Y\}$ is defined
by the equivalent formulas
$$\eqalign{
\II\{ X : Y\} &= \Eta\{X\} + \Eta\{Y\} - \Eta\{X,Y\} \cr
&= \Eta\{X\} - \Eta\{X\given Y\} \cr
&= \Eta\{Y\} - \Eta\{Y\given X\}. \cr
}$$
It measures, roughly speaking, how much information one can get from one variable
by looking at the other one. From the formula it is clear that $\II\{X : Y\}  = 0$
if $X$ and $Y$ are independent. In general, we still have the inequality $\II\{X : Y\} \ge 0$,
which a corollary of the following lemma, which expresses the intuitive fact that conditioning
cannot increase the entropy of a random variable.
The proof, which is due to C.~West, uses an argument similar to the one
we used to prove Proposition~{\propmaxprob}.

@@

The general case follows from the continuity axiom and the fact that any discrete random
variable, regarded as a vector in $l_1(\RR)$,
by vectors with finite support, and these in turn can be approximated by vectors of finite support
and rational coordinates.\slug

@@

\edef\cormaxsumineq{\the\sectcount.\the\thmcount}
\proclaim Corollary \advthm. If $X$ and $Y$ are independent, then
$$\max\bigl( \Eta\{X\}, \Eta\{Y\} \bigr) \le \Eta\{ X \pm Y\}.$$

@@

The entropic analogue of the Ruzsa distance is defined as follows.
For finitely supported random variables $X$ and $Y$ taking values in the same abelian group,
we let $X'$ and $Y'$ be independent copies of $X$
and $Y$, respectively, and define the {\it entropic Ruzsa distance} by
$$\dd\{X,Y\} = \Eta\{X'-Y'\} - {\Eta\{X'\} \over 2} - {\Eta\{Y'\}\over 2}.$$
This definition only depends on the individual distributions of $X$ and $Y$ and does not require them to
have the same sample space. When $X$ and $Y$ are independent, the Ruzsa distance enjoys some
nice properties.

\edef\propabsdist{\the\sectcount.\the\thmcount}
\proclaim Proposition \advthm. Let $X$ and $Y$ be independent taking values in the same group. Then
$$\bigl| \Eta\{X\} - \Eta\{Y\} \bigr| \le 2\dd\{X,Y\}$$
and
$$\max\bigl( \Eta\{X-Y\} - \Eta\{X\}, \Eta\{X-Y\} - \Eta\{Y\} \bigr) \le 2\dd\{X,Y\}$$

\proof By independence we have
$$2\dd\{X,Y\} = 2\Eta\{X-Y\} - \Eta\{X\} - \Eta\{Y\}\over 2,$$
Corollary~{\cormaxsumineq} tells us that $2\Eta\{X-Y\} \ge 2\Eta\{X\}$, so this is at least $\Eta\{X\}-\Eta\{Y\}$.
The same corollary also says that $2\Eta\{X-Y\} \ge 2\Eta\{Y\}$, which allows us to add the absolute value bars
to the inequality.

The second inequality is proved using the same corollary, but only applying it to one of the $\Eta\{X-Y\}$
terms.\slug

Similarly to the Ruzsa distance on sets, we don't necessarily have $\dd\{X,X\} = 0$, but we do have
the triangle inequality, which shall now prove.

@@

If $A$ and $B$ are subsets of the same abelian group such that $|A+B|$ is much smaller than $|A|\cdot|B|$,
then it stands to reason that there must be a lot of redundancy in $A+B$; that is, many elements of $A+B$ can
be expressed as $a+b$ in lots of different ways. To capture this notion, we can define the
{\it additive energy} between two sets $A$ and $B$ to be
$$E(A,B) = \bigl| \bigl\{ (a,a',b,b') \in A\times A\times B\times B : a + b = a'+b'\bigr\}\bigr|.$$

@@

\edef\propdoublingone{\the\sectcount.\the\thmcount}
\proclaim Proposition \advthm. Let $A$ be a finite subset of an abelian group $G$. If $|A+A| = |A|$,
then $A$ is a coset of a subgroup $H$.

@@

If $x,y\in H$, then $A+x = A$ and $A+y=A$, so subtracting $y$ from both sides of the second identity we obtain
$A-y = A$, and adding $x$ to both sides of this, we get
$$A + x-y = A+x = A,$$
so $x-y\in H$.
This along with the fact that $0\in H$ shows that $H$ is a subgroup of $G$.

@@

The proof of this theorem will occupy the
remainder of these notes. As a first step, we shall show that
Theorem~{\thmpfr} follows from an entirely information-theoretic statement.
In this and the rest of the notes, for $A$ finite
we use the notation $U_A$ to denote the random variable that is uniform on $A$.

\edef\thmsubgroup{\the\sectcount.\the\thmcount}
\proclaim Theorem \advthm. There is a constant $C'$ such that the following holds.
Let $G = \FF_2^n$ and let $X_1^*$ and $X_2^*$ be random variables taking values in
$G$. There is some subgroup $H\subseteq G$ such that
$$\dd\{ X_1^*, U_H\} + \dd\{ X_2^*, U_H\} \le C'\dd\{X_1^*, X_2^*\}.$$

In the proof that this theorem implies the previous one, we shall need the Ruzsa covering lemma,
whose proof is short enough that we include it for completeness.

\edef\lemruzsacovering{\the\sectcount.\the\thmcount}
\parenproclaim Lemma {\advthm} (Ruzsa covering lemma). Let $G$ be an abelian group and let $A$ and
$B$ be finite subsets of $Z$. Then there is a set $L\subseteq A$ of size at most $|A+B|/|B|$ such that
$A\subseteq L+B-B$.

\proof Let $L = \{a_1,\ldots,a_l\}$ be a maximal subset of $A$ with the property that the sets $a_i+B$
are disjoint. For all $a\in A$ there is $i$ such that $(a+B) \cap (a_i+B) \ne \emptyset$, otherwise we
could add $a$ to $L$ and contradict maximality. Equivalently, we can find $b$ and $b'$ such that
$a+b = a_i + b'$ so $a\in a_i + B - B$. Hence $A\subseteq L+B-B$. Lastly, $|L|\cdot |B| = |K+B| \le |A+B|$.\slug

We now show that if Theorem~{\thmsubgroup} holds for some constant $C'$, then
Theorem~{\thmpfr} holds with $C=C'+1$.

\medskip\noindent{\it Proof of Theorem~{\thmpfr} assuming Theorem~{\thmsubgroup}.}\enspace
Let $A\subseteq \FF_2$ and $K\ge 1$
such that $|A+A|\le K|A|$. Let $U_A$ be the uniform distribution on $A$, so that $\Eta\{U_A\} = \lg|A|$.
Letting $U_A'$ and $U_A''$ be two independent copies of $U_A$, the sum $U_A' - U_A''$ belongs to $|A+A|$,
so $\Eta\{U_A' + U_A''\} \le \lg|A+A|$, and we have
$$\dd\{U_A, U_A\} = \Eta\{U_A' + U_A''\} - \Eta\{U_A\} \le \lg|A+A|-\lg|A| \le \lg K,$$
since $U_A' + U_A'' = U_A' - U_A''$ in $\FF_2^n$. By Theorem~{\thmsubgroup}, we can find
$H\subseteq G$ such that
$$2\dd\{ U_A, U_H\} \le C'\lg K,$$
where $U_H$ is taken to be uniform on $H$ independently of $U_A$. Independence and the definition
of Ruzsa distance let us conclude
$$\Eta\{U_A-U_H\} \le {\lg|A| + \lg|H| + C'\lg K\over 2},$$
from which applying Proposition~{\propmaxprob} allows us to procure some $x_0 \in \FF_2^n$ such that
$$\pr\{ U_A - U_H = x_0\} \ge {1\over |A|^{1/2} |H|^{1/2} K^{C'/2}}.$$
Since $U_A$ and $U_H$ are uniformly and independently random on $A$ and $H$ respectively, we rewrite this as
$${\bigl| A\cap (H+x_0)\bigr| \over |A|\cdot|H|}  \ge {1\over |A|^{1/2} |H|^{1/2} K^{C'/2}},$$
whence
$$\bigl| A\cap (H+x_0)\bigr|  \ge {|A|^{1/2} |H|^{1/2} \over K^{C'/2}}.$$
By the Ruzsa covering lemma applied to the sets $A$ and $A\cap (H+x_0)$, there is a set $L\subseteq A$ of size
$$\eqalign{
|L| &\le {\bigl| A + (A\cap (H+x_0))\bigr|\over \bigl| A\cap (H+x_0)\bigr|} \cr
&\le {|A + A|\over \bigl| A\cap (H+x_0)\bigr|} \cr
&\le {K|A|\cdot K^{C'/2} \over |A|^{1/2} |H|^{1/2} } \cr
&= {K^{C'/2+1}|A|^{1/2} \over |H|^{1/2} } \cr
}$$
such that
$$A \subseteq L+\bigl(A\cap (H+x_0)\bigr) - \bigl(A\cap (H+x_0)\bigr) \subseteq L+H.$$
Proposition~{\propabsdist} tells us that
$$\bigl| \lg|H| - \lg|A|\bigr| = \bigl| \Eta\{U_H\} - \Eta\{H_A\}\bigr| \le 2\dd\{ U_A, U_H\} \le C'\lg K,$$
so
$$\max\biggl( {|H|\over |A|}, {|A|\over |H|}\biggr) \le K^{C'}$$
and we have
$$|L| \le K^{C'+1}.$$
If $|H|\le |A|$ then we are done (and the factor of $2$ is unnecessary in the theorem statement),
since $A$ is a subset of $|L|$ cosets of $H$.

If not, then we can pick a
subgroup $H'$ of $H$ with $|A|/2 \le |H'|\le |A|$ by iteratively taking hyperplanes until the condition is met.
The group $H$ may be covered by $|H|/|H'| \le 2|H|/|A|$ translates of $H'$, so $A$ can be covered by
$${2|H|\over |A|}|L| \le {2|H|\over |A|}\cdot {K^{C'/2+1}|A|^{1/2} \over |H|^{1/2} } 
= 2 {C'/2+1} {|H|^{1/2} \over |A|^{1/2}} \le 2K{C'+1}$$
translates of $H'$.\slug

\advsect A compactness argument

We have reduced the proof of the Freiman--Ruzsa theorem to proving the inform\-ation-theoretic
Theorem~{\thmsubgroup}. In this section, we shall reduce the proof to a different statement via a compactness
argument. First we recall some basic topological properties of probability measures.

Fix a finite set $G$ of size $n$. Ordering the elements of $G$ from $1$ to $n$, one can associate
to each probability distribution $\mu$ on $G$ the vector
$$\bigl( \mu(\{1\}), \mu(\{2\}), \ldots, \mu(\{n\})\bigr)$$
in $\RR^n$. (Starting now we will write $\mu_i$ instead of $\mu(\{i\})$.)
Let $\P(G)$ denote the set of all probability distributions on $G$.
This is a metric space under the {\it total variation
distance} $\dTV$, where if $\mu = (\mu_1,\ldots,\mu_n)$
and $\nu = (\nu_1,\ldots,\nu_n)$ are two probability distributions on $G$, then
$$\dTV(\mu,\nu) = \sum_{i=1}^n |\mu_i - \nu_i|.$$
Regarding the space as a subset of $\RR^n$, this norm is equivalent to the $L_1$ norm; hence
the total variation metric on $\P(G)$ is equivalent to the Euclidean metric on $\RR^n$.
Since probabilities are in $[0,1]$, the space $\P(G)$ is a subset of the closed set $[0,1]^n$,
and since $\P(G)$ is the inverse image of the set $\{1\}$ under the continuous function
$$(x_1,\ldots,x_n) \mapsto x_1 + \cdots + x_n,$$
we conclude that $\P(G)$ is a compact topological space. In passing, we note here that this is
the topology under which the continuity axiom asserts that $\Eta\{X\}$ is continuous.

Next we show that the Ruzsa distance is also continuous with respect to this topology.

\edef\propruzsacontinuous{\the\sectcount.\the\thmcount}
\proclaim Proposition \advthm. Let $G$ be a finite abelian group.
The Ruzsa distance $\dd$, regarded as a functional
from $\P(G)\times \P(G)\to \RR$, is continuous.

Let $X$ and $Y$ be random variables; without loss of generality, we may assume they are independent,
so that
$$\dd\{X,Y\} = \Eta\{X-Y\} - {\Eta\{X\} \over 2} - {\Eta\{Y\}\over 2}.$$
The entropy functional is continuous by axiom, so from here it suffices to show that
the function $f : \P(G)\times \P(G) \to \P(G)$ mapping $(X,Y)\mapsto X-Y$ is continuous.

Since all norms on finite-dimensional spaces are equivalent, we may work with
$\P(G)\times \P(G)$ and $\P(G)$ as normed spaces using the $L_1$ norm on $\RR^{2n}$ and $\RR^n$
respectively.
Let $\mu_g = \pr\{X = g\}$, $\nu_g = \pr\{ Y = g\}$, and $\xi_g = \pr\{X-Y = g\}$ for all $g\in G$.
Given another pair $(X', Y')$ of random variables, define $\mu_g'$, $\nu_g'$, and $\xi_g'$ accordingly.
Now let $\eps > 0$ and suppose that the vectors $(\mu, \nu)$ and $(\mu', \nu')$
have $\bignorm{(\mu,\nu), (\mu',\nu')}_1 < \delta$ for some $\delta>0$ to be specified later; writing
this out in full, we have
$$\sum_{g\in G} \big(|\mu_g - \mu_g'| + |\nu_g - \nu_g|\bigr) < \delta.$$
Observe that
$$\xi_g = \pr\{X-Y= g\} = \sum_{h\in G} \pr\{X = g+h\} \pr\{Y = h\} = \sum_{h\in G} \mu_{g+h}\nu_h$$
and similarly for $\xi_g'$. From this we can expand and bound
$$\eqalign{
\norm{\xi, \xi'}_1 &= \sum_{g\in G} |\xi_g - \xi_g'| \cr
&= \sum_{g\in G} \Bigl| \sum_{h\in G} \mu_{g+h}\nu_h - \sum_{h\in G} \mu_{g+h}\nu_h\Bigr| \cr
&\le \sum_{g\in G} \sum_{h\in G} \bigl( |\mu_{g+h}\nu_h - \mu_{g+h}'\nu_h| +
|\mu_{g+h}'\nu_h - \mu_{g+h}'\nu_h'|\bigl) \cr
&= \sum_{h\in G} \nu_h \sum_{g\in G} |\mu_{g+h} - \mu_{g+h}'|
+ \sum_{g\in G} \mu_g \sum_{h\in G} |\nu_h - \nu_h'| \cr
&= \sum_{g\in G} |\mu_{g} - \mu_{g}'| + \sum_{h\in G} |\nu_h - \nu_h'| \cr
&< \delta,
}$$
so in fact we can set $\delta = \eps$.\slug

So much for topology. Returning to entropies, we now prove an entropic version of
Proposition~{\propdoublingone}.

\edef\propentropydoublingone{\the\sectcount.\the\thmcount}
\proclaim Proposition \advthm. Let $X$ be a random variable taking values in an abelian group $G$.
Then $\dd\{X, -X\} = 0$ if and only if $X$ is uniformly distributed on a coset of a finite subgroup of $G$.

\proof If $X$ is the uniform distribution on a coset $a+H$ of a finite subgroup $H$, then letting $X'$ be an
independent copy of $X$, the random variable $X+X'$ is uniform on the coset $a+a+H$, which has the
same size as $|a+H|$, so $\Eta\{X+X'\} = \Eta\{X\}$, meaning that
$$\dd\{X, -X\} = \Eta\{X+X'\} - {\Eta\{X\}\over 2} - {\Eta\{X'\} \over 2} = 0.$$
On the other hand, suppose that $\Eta\{X+X'\} = \Eta\{X\}$, where $X'$ is an independent copy of $X$.
Then
$$\eqalign{
\Eta\{X+X'\} &= 2\Eta\{X\} - \Eta\{X\} \cr
&= \Eta\{X, X'\} - \Eta\{X'\} \cr
&= \Eta\{X+X', X'\} - \Eta\{X'\} \cr
&= \Eta\{X+X'\given X'\},\cr
}$$
so $X+X'$ is independent of $X'$.

Let $A$ denote the support of $X$. We have shown that the distribution
$(X + X' \given X' = x)$ is the same
regardless of our choice of $x\in A$. Then for all $x,y\in A$, the distribution of $X+x$ is the
same as that of $X+y$, so the distribution of
of $X+x-y$ is the same as that of $X$. This implies that $A$ is finite,
$X$ is the uniform distribution on $A$,
and adding an element $h\in A-A$ to the set $A$ simply permutes the set.

As in the proof
of Proposition~{\propdoublingone}, let $H = \{h\in G : A+h = A\}$. We proved back then that $H$ is a subgroup
of $G$, and the conclusion of the previous paragraph
asserts that $A-A \subseteq H$. But letting $h\in H$, the fact that
$h+A = A$ implies that $h\in A-A$. We conclude that
$A-A$ is a finite subgroup of $G$, whence $A$ is a coset of a finite subgroup of $G$.\slug

From this, we are able to prove the special case of Theorem~{\thmsubgroup} in which the Ruzsa distance between
the two given random variables is zero.

\edef\lemzerosubgroup{\the\sectcount.\the\thmcount}
\proclaim Lemma \advthm. Let $X_1$ and $X_2$ be random variables taking values in $\FF_2^n$ with
$\dd\{X_1,X_2\} = 0$. Then there exists a subgroup $H$ of $\FF_2^n$ such that
$$\dd\{X_1, U_H\} = \dd\{X_2,U_H\} = 0.$$

\proof The triangle inequality gives $\dd\{X_1,X_1\} = 0$, and since $-x = x$ in $\FF_2^n$,
we also have $\dd\{X_1, - X_1\} = 0$. By the previous proposition, there is a subgroup $H$ such that
$X_1$ has the same distribution as $H$. So $\dd\{X_1, U_H\} = 0$ and by the triangle inequality once again,
$\dd\{X_2, U_H\} = 0$ as well.\slug

Recall that in the previous section, we reduced the proof of the polynomial Freiman--Ruzsa theorem
to a statement involving some random variables $X_1^*$ and $X_2^*$. For the remainder of these notes,
we shall consider these variables to be fixed, and we also fix $\eta = 1/9$ for short.
Defining the functional
$$\tau(X_1, X_2) = \dd\{X_1, X_2\} - \eta\dd\{X_1^*, X_1\} - \eta\dd\{X_2^*, X_2\},$$
we can then reduce the proof of Theorem~{\thmsubgroup} (and consequently the entire proof) to showing the
following proposition.

\edef\propmaintau{\the\sectcount.\the\thmcount}
\proclaim Proposition \advthm. Let $X_1$ and $X_2$ be two $\FF_2^n$-valued random variables with
$\dd\{ X_1, X_2\} > 0$. Then there are $\FF_2^n$-valued random variables $X_1'$ and $X_2'$ such that
$$\tau\{X_1', X_2'\} < \tau\{X_1, X_2\}.$$

If we can prove this proposition, then we have Theorem~{\thmsubgroup} with a certain constant $C'$ which
we shall make explicit during the proof.

\medskip\noindent{\it Proof of Theorem~{\thmsubgroup} assuming Proposition~{\propmaintau}}.\enspace
We proved earlier that $\P(\FF_2^n)$ is a compact topological space, thus so its Cartesian product with
itself. By Proposition~{\propruzsacontinuous}, the Ruzsa distance functional is continuous on this space,
which implies that $\tau$ is as well. Hence $\tau$ attains its infimum on the space $\P(\FF_2^n)^2$;
let $X_1$ and $X_2$ be two distributions such that $\tau\{X_1, X_2\}$ is minimal. By the contrapositive
of Proposition~{\propmaintau}, we must have $\dd\{X_1, X_2\} = 0$, so by Lemma~{\lemzerosubgroup},
there exists a subgroup $H$ of $\FF_2^n$ such that
$$\dd\{X_1, U_H\} = \dd\{X_2, U_H\} = 0.$$
This means that
$$\dd\{X_1^*, U_H\} \le \dd\{X_1^*, X_1\} + \dd\{ X_1, U_H\} = \dd\{X_1^*, X_1\}$$
and
$$\dd\{X_1^*, X_1\} \le \dd\{X_1^*, U_H\} + \dd\{U_H, X_1\} = \dd\{X_1^*, U_H\},$$
so $\dd\{X_1^*, U_H\} = \dd\{X_1^*, X_1\}$. The same holds for $X_2^*$ and $X_2$.
Then since $\dd\{X_1, X_2\} = 0$ and by minimality of $\tau\{X_1, X_2\}$, we obtain
$$\eqalign{
\eta\dd\{X_1^*, U_H\} + \eta\dd\{X_2^*, U_H\} &= \eta\dd\{X_1, U_H\} + \eta\dd\{X_2, U_H\} \cr
&= \tau\{X_1, X_2\} \cr
&\le \tau\{X_1^*, X_2^*\} \cr
&= \dd\{X_2^*, X_1^*\} + \eta\dd\{X_1^*, X_2^*\} + \eta\dd\{X_2^*, X_1^*\} \cr
&= (1+2\eta) \dd\{X_1^*, X_2^*\}. \cr
}$$
From our choice of $\eta = 1/9$ we can multiply both sides by $9$ to get
$$\dd\{X_1^*, U_H\} + \dd\{X_2^*, U_H\} \le 11\dd\{X_1^*, X_2^*\},$$
and furthermore, since
$$\bigl| \dd\{X_1, U_H\} - \dd\{X_2, U_H\} \bigr| \le \dd\{X_1, X_2\}$$
by two invocations of the triangle inequality, neither $\dd\{X_1^*, U_H\}$ nor
$\dd\{X_2^*, U_H\}$ can be greater than $6\dd\{X_1^*, X_2^*\}$.\slug

For convenience, we restate what we have just proved with the constants plugged in.

\proclaim Theorem~{\thmsubgroup}$\mathbold'$.
Let $X_1^*$ and $X_2^*$ be random variables taking values in
$\FF_2^n$. There is some subgroup $H\subseteq \FF_2^n$ such that
$$\dd\{ X_1^*, U_H\} + \dd\{ X_2^*, U_H\} \le 11\dd\{X_1^*, X_2^*\},$$
and furthermore,
$$\max\bigl( \dd\{X_1^*, U_H\}, \dd\{X_2^*, U_H\}\bigr) \le 6\dd\{X_1^*, X_2^*\}.\noskipslug$$
